---
title: "Final Project Group 1 BUAN 6365"
date: "August 1, 2021"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_notebook:
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
  html_document:
    toc: yes
editor_options:
  chunk_output_type: inline
---

# Setup and Packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load pacman, warning=FALSE, message=FALSE, results='hide'}
if(!require("pacman")) install.packages("pacman")
pacman::p_load(here, knitr, rmarkdown, gplots, GGally, mosaic, scales, mlbench, 
               data.table, ggplot2, dplyr,ggplot, tidyr,tseries, sandwich, lmtest, 
               gridExtra, rpivotTable, pivottabler, reshape2,  mapproj,  caret, 
               ggpubr, forecast, leaps, tidyverse, MASS, esquisse, GGally, gganimate, 
               scales, corrplot, rattle, e1071, viridis, hrbrthemes, vcd, timeDate, car)

```

# Data 

### 1. Dataset : Exploring the Data

```{r Read CSV file}

Salary_data.df <- read.csv("Employee_Salary_Compensation.csv")

summary(Salary_data.df)

```

### 2. Grouping the Data for later use

```{r Data Grouping}
# Data Grouping
Employee.dt <- setDT(Salary_data.df)
Employee.year.group <- group_by(Salary_data.df, Year, Organization.Group)
Employee.org.group <- group_by(Salary_data.df, Organization.Group)
Employee.dept.group <- group_by(Employee.org.group, Department)
Employee.family.group <- group_by(Employee.dept.group, Job.Family)
Employee.Job.group <- group_by(Employee.family.group, Job)
```

### 3. Data Filtering

We are removing "Organization Group Code", "Job Family Code", "Job Code", "Year Type", 
"Department Code", "Union Code" etc. and other unnecessary columns from the data.

```{r Filter CSV File}

Salary1.df <- Salary_data.df[, -c(1:4,7,9,13)]

Salary1.dt <- setDT(Salary1.df)

str(Salary1.df)

```

### 4. Removing Null Values 

Here, we have removed the null values and removed the category "General City Responsibilities" 
from Organization Group. We have explained the category removal further in the EDA section.

```{r Removing Null Values}

Salary2.df <- subset(Salary_data.df, 
                     Organization.Group != "General City Responsibilities" & 
                       Salaries > 0 & Overtime >= 0 & Other.Salaries >= 0 & 
                       Total.Salary > 0 & Retirement >= 0 & 
                       Health.and.Dental >= 0 & Total.Benefits >= 0 & 
                       Total.Compensation >= 7.25)

str(Salary2.df)

```


# Explorartory Data Analysis

### 1. Correlation

Heat Map Representing Various Salaries

```{r HeatMap}

sal.df <- Salary_data.df[,-c(1:13, 17, 21)]

heatmap.2(cor(sal.df), Rowv = FALSE, Colv = FALSE, dendrogram =
"none",cellnote = round(cor(sal.df),2),notecol = "black", key = FALSE, 
trace = 'none', margins = c(10,10))
```

### 2. Univariate data analysis 

Here we are visually exploring each variable to understand the various characteristic and the data spread.

```{r}
options(scipen=999)
d_sal <- density(Salary_data.df$Salaries) 
plot(d_sal) 

d_overtime <- density(Salary_data.df$Overtime, xName='Overtime') 
plot(d_overtime) 

d_othersal <- density(Salary_data.df$Other.Salaries, xName='Other Salaries') 
plot(d_othersal) 

d_totsal <- density(Salary_data.df$Total.Salary, xName='Total Salary') 
plot(d_totsal) 

d_retire <- density(Salary_data.df$Retirement, xName='Retirement') 
plot(d_retire) 

d_health <- density(Salary_data.df$Health.and.Dental, xName='Health and Dental') 
plot(d_health) 

d_otherbene <- density(Salary_data.df$Other.Benefits, xName='Other Benefits') 
plot(d_otherbene) 

d_totbene <- density(Salary_data.df$Total.Benefits, xName='Total Benefits') 
plot(d_totbene) 

d_totcomp <- density(Salary_data.df$Total.Compensation, xName='Total Compensation') 
plot(d_totcomp)
polygon(d_totcomp, col="red", border="blue")

```


### 3. Numerical independent variable

Here we are visually exploring the numerical independent variable and how these 
variables affect the Total Compensation. (Bi-variate analysis)


```{r scatter plot of Total.Compensation with salaries, other salaries, retirement, health and dental, other benefits} 

options(scipen=999)
emp_sal_data.df <- Salary_data.df

ggplot(emp_sal_data.df, aes(x = Salaries, y = log(Total.Compensation))) +
  geom_point(color = "tomato2", alpha = 0.4)+
  xlab("Salaries") + 
  ylab("Total Compensation ($)")

ggplot(emp_sal_data.df, aes(x = Overtime, y = log(Total.Compensation))) +
  geom_point(color = "tomato2", alpha = 0.4)+
  xlab("Overtime") + 
  ylab("Total Compensation ($)")

ggplot(emp_sal_data.df, aes(x = Other.Salaries, y = log(Total.Compensation))) +
  geom_point(color = "tomato2", alpha = 0.4)+
  xlab("Other Salaries") + 
  ylab("Total Compensation ($)")

ggplot(emp_sal_data.df, aes(x = Retirement, y = log(Total.Compensation))) +
  geom_point(color = "tomato2", alpha = 0.4)+
  xlab("Retirement") + 
  ylab("Total Compensation ($)")

ggplot(emp_sal_data.df, aes(x = Health.and.Dental, y = log(Total.Compensation))) +
  geom_point(color = "tomato2", alpha = 0.4)+
  xlab("Health and Dental") + 
  ylab("Total Compensation ($)")

ggplot(emp_sal_data.df, aes(x = Other.Benefits, y = log(Total.Compensation))) +
  geom_point(color = "tomato2", alpha = 0.4)+
  xlab("Other Benefits") + 
  ylab("Total Compensation ($)")

```


### 4. Total Compensation V/S Year

The graphs in 4 & 5 show us why we needed to remove the category "General City Responsibilities".
We cannot see any data from "General City Responsibilities" in graph 4 in any year. This  indicated that we needed to dig deeper to analyze what was going wrong. In graph 5,
we can see that there is a big dip in the average salaries throughout the years for this
category. We can safely assume that there is data missing from this category and for a more accurate
model, we decided to remove the category.

```{r Average Total Compensation (in Million)}

options(scipen=999)
Comp_Year <- ggplot(Salary1.df, 
           aes(x=Year, y=mean(Total.Compensation), fill=Organization.Group, 
               width=0.5 )) +
  geom_bar(stat="identity") + 
  ggtitle("Year over Year Average Total Compensation") +
   scale_y_continuous(
        name = "Average Total Compensation (per Million $)",
        labels = function(y) paste0(y / 1000000, "M")) +
  theme_classic()

Comp_Year
```

### 5. Year over Year Average Salary 


```{r Year over Year Average Salary}

ggplot(year.org, aes(x = Year, y = mean.salary, group = Organization.Group, 
                     color = Organization.Group)) +
  geom_line() +ggtitle("Year Over Year Average Salary") +
  xlab("Year") +ylab("Average Salary Amount ($)")
```

### 6. Top 5 Departments with highest Total Compensation

```{r Top 5 Departments with highest Total Compensation}

top15_totalcomp <- Salary1.df %>% 
arrange(desc(Total.Compensation)) %>% 
slice_max( Total.Compensation, n = 15)

top15_totalcomp

# Barplot displaying Top 5 departments with highest total compensation
ggplot(top15_totalcomp, aes(x=forcats::fct_rev(Department), y=Total.Compensation, width = 0.6 )) + 
    geom_bar(stat = "identity", fill = "#0096FF") +
  scale_x_discrete (
    name = "Department"
  ) +
  scale_y_continuous(
  name = "Total Compensation (per thousand $)",
  labels = function(y) paste0(y / 1000, "k")) +
  ggtitle("Top 5 Departments w.r.t Total Compensation ")

```

### 7. Top 5 Organization Group with highest Total Compensation

```{r Top 5 Organization Group with highest Total Compensation}

top350_totalcomp <- Salary1.df %>% 
arrange(desc(Total.Compensation)) %>% 
slice_max( Total.Compensation, n = 350)

top350_totalcomp

# Barplot Top 5 Organization Group with highest total compensation
ggplot(top350_totalcomp, aes(fct_rev(fct_reorder(Organization.Group,Total.Compensation)), Total.Compensation, width=0.3  )) + 
  geom_bar(stat = "identity", fill = "#FDDA0D") + 
  scale_x_discrete (
    name = "Organization Group"
  ) +
  scale_y_continuous(
  name = "Total Compensation (per Million $)",
  labels = function(y) paste0(y / 1000000, "M")) +
  ggtitle("Top 5 Organization Groups w.r.t Total Compensation")+
  theme(axis.text.x=element_text(angle=45, hjust=1))

```

### 8. Total Salaries by Year, Top 5 Departments, and  Organization Group

```{r Total Salaries by Year, Top 5 Departments, and  Organization Group}


ggplot(data=top15_totalcomp, 
       aes(x=factor(Year), y=Total.Compensation, 
           group=Organization.Group,
           color=Organization.Group)) + 
               geom_line() + 
               geom_point() +
               scale_x_discrete("Year") +
               scale_y_continuous(
                  name = "Total Compensation (per thousand $)",
                  labels = function(y) paste0(y / 1000, "K")) + 
          rotate_x_text() +
          ggtitle("Total Compensation by Year, Top 5 Departments, and  Organization Group") +
          facet_wrap(.~Department )
```

### 9. Yearly distribution of count of records w.r.t. Top 5 Organization Group

```{r Yearly distribution of count of records w.r.t. Top 5 Organization Group}

ggplot(data = top350_totalcomp, aes(x = Year, fill = Organization.Group)) +
    geom_bar() +
    scale_y_continuous(
      name = "Number of Records"
    ) +
  ggtitle("Yearly distribution of count of records w.r.t. Top 5 Organization Group")


```


# Regression Models

### Data Partition - 1 

```{r Data partition}
# Assigning 80% of data to training dataset and 20% of data to validation dataset.

Salary.df <- Salary_data.df[, -c(1:4,9,13)]
Salary_newSet.df <- Salary.df[, c(1, 8:10,12:14,16)]

set.seed(42)

training_dataset.index<- sample(c(1:nrow(Salary_newSet.df)), 
                                round(0.8 *nrow(Salary_newSet.df),0))

training_dataset.df<-Salary_newSet.df[training_dataset.index,]

validation_dataset.df<-Salary_newSet.df[-training_dataset.index,]

```

### Multicollinear Model

Because this R^2 value is too high, which is close to 1. Also, the Salaries and Retirement Health.and.Dental VIF value is greater than 10. So this model is multi-collinear, we can not run the linear regression for this model

```{r build check multicollinear model}
# Build the model
Salary.lm <- lm(Total.Compensation ~ ., data = training_dataset.df)

# Make predictions
predictions <- Salary.lm %>% predict(validation_dataset.df)
# Model performance
data.frame(
  RMSE = RMSE(predictions, validation_dataset.df$Total.Compensation),
  R2 = R2(predictions, validation_dataset.df$Total.Compensation)
)
print("summary of this linear model")
summary(Salary.lm)
print("VIF")
vif(Salary.lm)

```

### Data Partition - 2

```{r Data Partition}

quant <- quantile(Salary2.df$Total.Compensation, probs = seq(0, 1, .1))

Salary2.df <- Salary2.df %>% mutate(Seniority = case_when(
                                        Total.Compensation <= quant[4] ~ "Entry Level", 
                                        Total.Compensation  >= quant[4] & Total.Compensation  <= quant[6] ~ "Lower", 
                                        Total.Compensation  >= quant[6] & Total.Compensation  <= quant[8] ~ "Middle", 
                                        Total.Compensation  >= quant[8] & Total.Compensation  <= quant[10] ~ "Upper", 
                                        Total.Compensation  >= quant[10] ~ "Executive"))

set.seed(80)
train.index <- createDataPartition(Salary2.df$Total.Compensation, p = 0.8, list = FALSE)
train.df <- Salary2.df[train.index, ]
test.df <- Salary2.df[-train.index, ]

```

### Regression Analysis 1 Organization Group 

```{r Regression Analysis 1}


options(scipen = 999)
lm1 <- lm(Total.Compensation ~ Seniority, data = train.df)
summary(lm1)

lm2 <- lm(Total.Compensation ~ Seniority + Organization.Group, data = train.df)
summary(lm2)

lm3 <- lm(Total.Compensation ~ Seniority + Organization.Group + Seniority*Organization.Group, data = train.df)
summary(lm3)

par(mfrow = c(2,2))
plot(lm3)

lm_predict <- predict(lm3, test.df)
accuracy(lm_predict, test.df$Total.Compensation)

```

### Regression Analysis 2 Job Family

```{r Regression Analysis 2}

lm4 <- lm(Total.Compensation ~ Seniority + Job.Family, data = train.df)
summary(lm4)

lm5 <- lm(Total.Compensation ~ Seniority + Job.Family + Seniority*Job.Family, data = train.df)
summary(lm5)

par(mfrow = c(2,2))
plot(lm5)

lm_predict2 <- predict(lm5, test.df)
accuracy(lm_predict2, test.df$Total.Compensation)

```


# Appendix

### Total Benefits V/S Year

```{r Total Benefits V/S Year Stacked Bar Chart}

e <-ggplot(data=Salary1.df, 
           aes(x=Year, y=mean(Total.Benefits), fill=Organization.Group)) +
  ggtitle("Year over Year Average Total Benefits") +
  xlab("Year") +
  ylab("Average Total Benfits Amount ($)") +
  geom_bar(stat="identity", width=0.5) + 
  theme_classic()

e
```

### Top 5 Departments with highest Total Salary

```{r Top 5 Departments with highest Total Salary}

top20_totalsal <- Salary1.df %>% 
arrange(desc(Total.Salary)) %>% 
slice_max( Total.Salary, n = 20)

top20_totalsal

# Barplot for displaying Top 5 departments with highest Total salary
ggplot(top20_totalsal, aes(x=forcats::fct_rev(Department), y= Total.Salary, width=0.5 )) + 
  geom_bar(stat = "identity", fill = "#FFA500") +
  scale_x_discrete (
    name = "Department"
  ) +
  scale_y_continuous(
        name = "Total Salary (per thousand $)",
        labels = function(y) paste0(y / 1000, "k")) +
   ggtitle("Top 5 Departments w.r.t Total Salary ")

```

### Top 5 Organization Group with highest Total Salary

```{r Top 5 Organization Group with highest Total Salary}

top350_totalsal <- Salary1.df %>% 
arrange(desc(Total.Salary)) %>% 
slice_max( Total.Salary, n = 350)

top350_totalsal

#Barplot displaying Top 5 Organization Group with highest Total Salary

ggplot(top350_totalsal, aes(fct_rev(fct_reorder(Organization.Group,Total.Salary)), Total.Salary, width=0.4 )) + 
  geom_bar(stat = "identity", fill = "#7cedff") +
  scale_x_discrete (name = "Organization Group") +
  scale_y_continuous(
        name = "Total Salary (per thousand $)",
        labels = function(y) paste0(y / 1000, "k")) +
   ggtitle("Top 5 Organization Group w.r.t Total Salary ")+ 
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Top 4 Departments with highest Total Benefits

```{r Top 4 Departments with highest Total Benefits}

top15_totalbenefits <- Salary1.df %>% 
arrange(desc(Total.Benefits)) %>% 
slice_max( Total.Benefits, n = 15)

top15_totalbenefits

# Barplot for displaying Top 4 departments with highest total compensation

ggplot(top15_totalbenefits, aes(x=forcats::fct_rev(Department), y=Total.Benefits, width = 0.6 )) + 
    geom_bar(stat = "identity", fill = "#0096FF") + scale_x_discrete (name = "Department")+ scale_y_continuous(name = "Total Benefits (per thousand $)",
  labels = function(y) paste0(y / 1000, "k")) +
  ggtitle("Top 4 Departments w.r.t Total Benefits ") + theme(axis.text.x=element_text(angle=45, hjust=1))

```

### Top 2 Organization Group with highest Total Benefits

```{r Top 2 Organization Group with highest Total Benefits}

# Barplot for displaying Top 2 Organization Group with highest total benefit

ggplot(top15_totalbenefits, aes(x=Organization.Group, y=Total.Benefits, width=0.3  )) + 
  geom_bar(stat = "identity", fill = "#FDDA0D") +
  scale_x_discrete (
    name = "Organization Group"
  ) +
  scale_y_continuous(
  name = "Total Benefits (per thousand $)",
  labels = function(y) paste0(y / 1000, "k")) +
  ggtitle("Top 2 Organization Groups w.r.t Total Benefits ")

```

### Top 10 Total Compensation by Year, Department, and  Organization Group

```{r Top 10 Total Compensation by Year, Department, and  Organization Group}

top10_totalcomp <- Salary1.df %>% 
arrange(desc(Total.Compensation)) %>% 
slice_max( Total.Compensation, n = 10)

top10_totalcomp

ggplot(data=top10_totalcomp, 
       aes(x=factor(Year), y=Total.Compensation, 
           group=Organization.Group,
           color=Organization.Group)) + 
               geom_line() + 
               geom_point() +
               scale_x_discrete("Year") +
               scale_y_continuous(
                  name = "Total Compensation (per thousand $)",
                  labels = function(y) paste0(y / 1000, "K")) + 
          ggtitle("Top 10 Total Compensation by Year, Department, and  Organization Group") +
           facet_grid(.~Department ) 
```

### Year over Year Average Benefits 

```{r Year over Year Average Benefits}
ggplot(year2.org, aes(x = Year, y = mean.benefit, group = Organization.Group, 
                      color = Organization.Group)) +
  geom_line() + ggtitle("Year Over Year Average Benefits") +
  xlab("Year") + ylab("Benefit Amount ($)")
```

### Overtime V/S Year

```{r Overtime V/S Year Stacked Bar Chart}
g <-ggplot(data=Salary1.df, 
           aes(x=Year, y=mean(Overtime), fill=Organization.Group)) +
  ggtitle("Year over Year Average Overtime Salary") +
  xlab("Year") +
  ylab("Average Overtime Salary ($)") +
  geom_bar(stat="identity", width=0.5) + 
  theme_classic()

g

```

### Retirement V/S Year

```{r Retirement V/S Year Stacked Bar Chart}
i <-ggplot(data=Salary1.df, 
           aes(x=Year, y=mean(Retirement), fill=Organization.Group)) +
  ggtitle("Year over Year Average Retirement Benefits") +
  xlab("Year") +
  ylab("Average Retirement Benefit Amount ($)") +
  geom_bar(stat="identity", width=0.5) + 
  theme_classic()

i
```

### Yearly distribution of count of records w.r.t. Top 5 Departments

```{r Yearly distribution of count of records w.r.t. Top 5 Departments}

ggplot(data = top15_totalcomp, aes(x = Year, fill = Department)) +
    geom_bar() +
    scale_y_continuous(
      name = "Number of Records"
    ) +
  ggtitle("Yearly distribution of count of records w.r.t. Top 5 Departments")


```

### Stepwise Regression using leaps Package

```{r Stepwise regression using leaps package}

Salary.lm <- lm(Total.Compensation ~ ., data = training_dataset.df)
Salary_stepwiseReg <-  regsubsets(Total.Compensation ~ ., data = training_dataset.df, 
                                 nbest = 1, nvmax = dim(training_dataset.df)[2], 
                                 method = "seqrep")

summary(Salary_stepwiseReg)$which

print("Adjusted R-squared Values:")
summary(Salary_stepwiseReg)$adjr2

print("Mallow's Cp Values:")
summary(Salary_stepwiseReg)$cp

```

### Backward selection model using stepAIC() function

```{r Backward selection model using stepAIC() function}


Salary.back.lm <- lm(Total.Compensation ~ ., data = training_dataset.df)

back.Salary.plane <- stepAIC(Salary.back.lm, direction = "backward")

Salary.back.lm.pred <- predict(Salary.back.lm, validation_dataset.df)

print("Backward Step AIC Model")
accuracy(Salary.back.lm.pred, validation_dataset.df$Total.Compensation)

```

### Forward selection model using step AIC() function

```{r Forward selection model using step AIC() function}
Salary.lm <- lm(Total.Compensation ~ ., data = training_dataset.df)
Salary.lm.null <- lm(Total.Compensation~1, data = training_dataset.df)

# use step() to run forward regression.
Salary.lm.fselect <- step(Salary.lm.null, scope=list(lower=Salary.lm.null, upper=Salary.lm), direction = "forward")
summary(Salary.lm.fselect)  # Which variables were added?
Salary.lm.fselect.pred <- predict(Salary.lm.fselect, validation_dataset.df)
print("Forward Step AIC Model")
accuracy(Salary.lm.fselect.pred, validation_dataset.df$Total.Compensation)

```


















